---
title: "Functional programming in R"
subtitle: "With the purrr package"
author: "Derek Hansen"
date: "9/18/2019"
output:
  ioslides_presentation:
    highlight: espresso
    css: styles.css
    mathjax: "default"
    keep_md: true 
  beamer_presentation:
    highlight: espresso
  slidy_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(purrr)
library(magrittr)
```

## What is Functional Programming?

<blockquote>
"Functional programming is a programming paradigm ... that treats computation as the evaluation of mathematical functions and avoids changing-state and mutable data."
- Wikipedia
</blockquote>

- Functions are like those in mathematics: 
   1.  They always return the same output for a given input
   2.  No side-effects (e.g. modification of global variables)
   3.  Functions that satisfy (1-2) are called **pure**
- Functions are **first-class** objects which can be passed as arguments to other functions (a.k.a. **functionals**)
- There is no changing state as the program runs; values are assigned once as functions of other values and inputs
- Everyday example: Microsoft Excel (without any VBA scripting!)

## Why use functional programming?

- Through avoiding mutable state and composing pure functions, an FP program is also a pure function of its input
- This makes FP programs...
  -  Modular
  -  Predictable
  -  Easier to test
  -  Avoid common pitfalls involving changing state (e.g. global variables)


## Why use functional programming?

Natural functionals in the FP paradigm include ...

-  **map**: (where $f: X \to Y$)
\[((x_1, ..., x_n),  f) \to (f(x_1), ..., f(x_n)))\]
-  **filter**: (where $f$ is a **predicate** function $f:X \to \{0, 1\}$)
\[((x_1, ..., x_n), f) \to (x_i : f(x_i) = 1)\]
-  **reduce**: (where $f$ is an **operator** function $f:X \times X \to X$)
\[((x_1, ..., x_n), f) \to f(x_1, f(x_2, f(x_3, f(...))))\]

## Functional Programming in R

- R is **multi-paradigm**: it does not strictly adhere to FP principles, but it offers capability to use FP patterns
- Examples in base R include: 
  -  ```Map```, ```lapply```, ```sapply```, ```apply```, ```vapply```, ```mapply```
  - ```Reduce```
  - ```Filter```
- The ```purrr``` package by Hadley Wickham et al improves the the functional programming tools to R which are **syntactically consistent** and **type-safe**.


## Mapping

- ```map``` is pretty much equivalent to ```lapply```, but has some additional features

```{r echo=TRUE}
library(purrr)
my_sqrt <- function(x) sqrt(x)
str(map(c(1,2,3,4,5), my_sqrt))
str(lapply(c(1,2,3,4,5), my_sqrt))
```

## Mapping

- If we want an atomic double vector instead of a list, the ```map_dbl``` ensures we always receive that.
- ```sapply``` does the same thing in this particular instance, but we can run into problems...

```{r echo=TRUE}
str(map_dbl(c(1,2,3,4,5), my_sqrt))
str(sapply(c(1,2,3,4,5), my_sqrt))
```


## Problem: sapply is not type-safe!

- Example: Our colleague worked hard to make ```my_sqrt``` handle any real number.
- They even overwrote the function ```my_sqrt``` to make the transition seamless!

```{r echo=TRUE}
sqrt_general <- function(x) {
  if(x >= 0) sqrt(x)
  else return(paste0(sqrt(abs(x)), "i"))
}

my_sqrt <- sqrt_general
my_sqrt(5)
my_sqrt(-5)

```

## Problem: sapply is not type-safe!

```{r echo=TRUE}
str(sapply(c(1,2,3,4,5), my_sqrt))
str(sapply(c(-1,2,-3,4,5), my_sqrt))
```

- This is a great way to propogate errors. We have no way to guarentee whether sapply will return a "double" vector or a "string" vector.

## map_dbl is type-safe!

```{r echo=TRUE}
str(map_dbl(c(1,2,3,4,5), my_sqrt))
try(str(map_dbl(c(-1,2,-3,4,5), my_sqrt)))
```

- The ```map_*``` family of functions allows us to explictly impose which type we expect the output vector to be.
  - They "return an atomic vector of the indicated type (or die trying)" (documentation)

## map_chr is type-safe!

```{r echo=TRUE}
map_chr(c(1,2,3,4,5), my_sqrt)
map_chr(c(-1,2,-3,4,5), my_sqrt)
```

- Suppose our colleague convinced our team lead that we should work exclusively with strings to accomodate complex numbers
- We use ```map_chr``` to reflect that now we want the output to be a character vector.
- No errors now because both doubles and characters can be coerced to double.

## map_* is type-safe!

-  ```sapply``` implicitly coerces to an atomic vector in the most general unit in the output for "convenience", but this is very prone to unexpected errors.
- Most of the time, it is better to be explicit to catch any errors early and keep type stability.
- Can also use ``_lgl`` for logical, ```_int``` for integer, ```_raw``` for raw type, ```_dfr``` and ```_dfc``` for data-table columns and rows.

## Some more cool features of map - anonymous functions

- Can construct function in the argument using symbol notation

```{r, echo=TRUE}
map_dbl(c(1,2,3,4,5), ~.x^2 + .x + sin(.x))
```

## Some more cool features of map - multiple arguments

<!-- - Can map multiple arguments.  -->
- Can use ```map2_*``` for 2 argument functions; ```pmap_*``` for n-argument functions
- The $i$th positional argument can be referenced with ```..i``` syntax.

```{r, echo=TRUE}
map2_dbl(c(1,2,3,4,5), c(5,6,8,9,11), ~.x^2 + .y^2 + sin(.x))
pmap_dbl(list(1:5, 11:15, 21:25), ~..1 + ..2 + ..3)
pmap_dbl(list(1:5, 11:15, 21:25), function(x,y,z) x+y+z)
```

## Some more cool features of map - imap

- Can use ```imap``` if the names of the input list/vector are important.
- ```imap_*(x, f(x,y))``` is equivalent to ```map2_*(x, names(x), f(x,y))```
- The type ```dfr``` indicates that we expect the function to output a **D**ata**F**rame **R**ow, which are then bound row-wise into a single dataframe.

```{r, echo=TRUE}
library(dplyr)
midterm_grades <- c(Dan = 100, Derek = 20, Rob = 100)
grade_tbl      <- imap_dfr(midterm_grades, ~tibble(name = .y, grade = .x, pass = .x >= 50))
grade_tbl
```

## Some more cool features of map - map_if

- ```map_if``` allows for use of a predicate function (or a vector) to only apply to certain values. 
- It always returns a list (since the input and output could be of different types).

```{r echo=TRUE}
str(map_if(midterm_grades, !grade_tbl$pass, ~NA_real_))
str(map_if(midterm_grades, ~.x <= 50, ~"FAIL!!"))
```

## Some more cool features of map - map_if

- ```modify_if``` is the same as ```map_if```, but enforces that the type is the same as the input

```{r echo=T}
str(modify_if(midterm_grades, ~.x <= 50, ~NA_real_))
```
```{r echo=T}
try(str(modify_if(midterm_grades, ~.x <= 50, ~"FAIL!!")))
```

## keep and discard

```{r, echo=TRUE}
## Only keep students who passed
keep(midterm_grades, ~.x >= 50)
## Remove students who passed to get a list of students on notice
discard(midterm_grades, grade_tbl$pass)
```

## purrr in the wild - succinctly extract results from different models


```{r, eval=FALSE, echo=TRUE}
library(dplyr)
library(magrittr)
aic_bic_tbl <- list(
  `Binary Poverty Indicator Interaction` = logis_res_census_binpoor,
  `Poverty Rate Interaction` = logis_res_census,
  `Income Interaction` = logis_res_census_inc_interact,
  `No Income` = logis_res_census_noincome,
  `No Poverty Rate` = logis_res_census_nopoor
) %>%
  map2_dfr(names(.), ~tibble(model = .y, aic = AIC(.x), bic = BIC(.x))) %>% 
  arrange(aic)
aic_bic_tbl
```

- Example directly from my applied qual. (Could have used ```imap_dfr```!)
<!-- - ```map2_df``` is equivalent to ```map2_dfr```, but may be deprecated in the future -->
- ```purrr``` was designed by the same authors as ```dplyr``` and plays nicely with other ```tidyverse``` functions (including the pipe object ```%>%```).

## purrr in the wild - reduce to best model
```{r, eval=FALSE, echo=TRUE}
best_model <- list(
  `Binary Poverty Indicator Interaction` = logis_res_census_binpoor,
  `Poverty Rate Interaction` = logis_res_census,
  `Income Interaction` = logis_res_census_inc_interact,
  `No Income` = logis_res_census_noincome,
  `No Poverty Rate` = logis_res_census_nopoor
) %>%
  reduce(~ifelse(BIC(.x) < BIC(.y), .y, .x))
```

- ```reduce``` function applies an operator function to reduce a vector to one value
- Illustrating example, but in reality it would be more efficient to use ```which.max(aic_bic_tbl$bic)``` (because it uses C code and more efficient algorithm)

## Conclusions

-  Through the Functional Programming (FP) paradigm, ```purrr``` allows for more concise and error-robust R coding patterns
-  Allows complex operations to be composed from simple building blocks by operating on user-specified functions
-  Many, many more features are contained in ```purrr``` beyond what was shown today

### Further reading

-  Tidyverse website (https://purrr.tidyverse.org/)
- "Iteration" chapter in R for Data Science (https://r4ds.had.co.nz/iteration.html)
-  Hadley's ```plyr``` package which handles array and data.frame inputs.
   <!-- -  Functionality-wise, ```plyr >= dplyr + purrr```  -->

# Thank You!



